Generating Verified LLVM from Isabelle/HOL

### 2019-06-07 paper/main.tex
Major restructuring, added thys-2019

\documentclass[a4paper,oribibl,envcountsame]{llncs}
\pagestyle{plain}
\usepackage[T1]{fontenc}
\usepackage[misc]{ifsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{bussproofs}
\usepackage{cite}
\usepackage{color}
\usepackage{enumitem}
%\let\vec=\undefined
%\usepackage{mdsymbol}  %%% indirectly needs texlive-generic-extra
\usepackage{stmaryrd}
\usepackage{subfig}
\usepackage{textcomp}
\usepackage{url}
\usepackage{xspace}
\usepackage{mathtools}
\usepackage{todonotes}
\usepackage{enumitem}

@@

\usepackage{array}
\newcolumntype{x}[1]{>{\centering\arraybackslash\hspace{0pt}}b{#1}}
\usepackage{diagbox}
\usepackage{siunitx}

@@

\usepackage[final,tracking=true,kerning=true,spacing=true,stretch=10,shrink=10]{microtype}

%% BEGIN Times font
\usepackage{mathptmx}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
%% END Times font

@@

\usepackage[
   a4paper,
   pdftex,
   pdftitle={Isabelle LLVM},
   pdfauthor={Peter Lammich},
   pdfkeywords={},
   pdfborder={0 0 0},
   draft=false,
   bookmarksnumbered,
   bookmarks,
   bookmarksdepth=2,
   bookmarksopenlevel=2,
   bookmarksopen]{hyperref}

\hyphenation{Isa-belle Nieuw-en-huis meta-logic multi-set multi-sets}

@@

@@

\title{Isabelle LLVM}
\titlerunning{Isabelle LLVM}

\author{Peter Lammich}
\authorrunning {P. Lammich}
\institute{
  Technische Universit\"at M\"unchen, Munich, Germany \\
  \email{lammich@in.tum.de}
}

@@

\begin{abstract}
We formalize the semantics of a fragment of the LLVM intermediate representation in Isabelle/HOL.
On top of this semantics, we have developed
\begin{enumerate}
\item a code generator, to translate from Isabelle/HOL to actual LLVM text; 
\item a verification condition generator, to reason about LLVM programs;
\item a backend for the Isabelle Refinement Framework, to enable a high-level refinement based approach to developing verified LLVM programs.
\end{enumerate}
As an example, we adopt an existing formalization of the Knuth-Morris-Pratt string search algorithm to our new framework.
With only slight changes to the original formalization, we obtain a verified LLVM implementation, which runs faster and uses less memory 
than the originally verified Standard ML implementation.

The trusted code base of our approach is the formalization of the semantics, which we have engineered for conciseness and readability, 
and the code generator that translates from Isabelle to LLVM text, which is merely a pretty printer. 
The other tools that we developed cannot compromise the soundness of the approach, as they only generate theorems that are proved via Isabelle's kernel.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
The Isabelle Refinement Framework~\cite{LaTu12,La13} features a stepwise refinement approach to verified algorithms, using the Isabelle/HOL theorem prover~\cite{NPW02,NiKl14}.
It has been successfully applied to verify many algorithms and software systems, 
among them LTL and timed automata model checkers\cite{ELNN13,BrLa18,WiLa18}, network flow algorithms~\cite{LaSe17,LaSe16}, a verified SAT-solver certification tool~\cite{La17_CADE,La17_SAT}, and even a verified SAT solver~\cite{FBL18}.
Using Isabelle/HOL's code generator~\cite{HKKN13}, the verified algorithms can be extracted to functional languages like Haskell or Standard ML.
However, the code generator only provides partial correctness guarantees, i.e., termination of the generated code cannot be proved.
Moreover, the generated code is typically much slower than the same algorithms implemented in C or Java.

The original Refinement Framework could only generate purely functional code. 
The first remedy to the performance problem was to introduce array data structures that behave like a functional 
list on the surface, but are implemented by a destructively updated array behind the scenes, similar to Haskell's now deprecated DiffArray.
While this gained some performance, the array-based implementation itself was not verified, such that we had to trust its correctness. 
Moreover, an array access still required a significant amount of overhead compared to a simple pointer dereference in C.

The next step was the Sepref tool~\cite{La15}. It generates code for Imperative HOL~\cite{BKHEM08}, which provides a heap monad inside Isabelle/HOL, 
and a code generator extension to generate code that uses the stateful arrays provided by ML, or the heap monad of Haskell.
The Sepref tool performs automatic data refinement from abstract data types like maps or sets to concrete implementations like hash tables, which 
can be placed on the heap and destructively updated. 
Moreover, it provides tools~\cite{La16} to assist in the definition of new data structures, exploiting `free theorems'~\cite{Wad89} that it 
obtains from parametricity properties of the abstract data types.
%
Using Imperative/HOL as backend, we gained some additional performance: For example, the GRAT tool~\cite{La17_SAT} provides a 
verified checker for UNSAT certificates in the DRAT format~\cite{WHH14}. It is faster than the unverified state-of-the art 
checker \textsc{drat-trim}~\cite{WHH14}, which is written in C. However, the GRAT tool spends most of its run time in an unverified 
certificate preprocessor. Nevertheless, optimizing the verified part of the code is important: The very same technique was also implemented in Coq,
using purely functional data structures~\cite{CMS17,CHHKS17}. There, the verified code was actually the 
bottleneck\footnote{Later, the checker was rewritten in ACL2, also using imperative data structures~\cite{CHHKS17,HHKW17}}.

This paper presents a next step towards efficient verified software: A refinement framework to generate verified code in LLVM intermediate representation~\cite{LLVM-manual} 
with total correctness guarantees. LLVM is an imperative intermediate language with a powerful and well-tested optimizing compiler.
We first formalize the semantics of a fragment\footnote{We support integer instructions, pointers, a simplified memory model, and simplified control flow. We expect this fragment to be sufficient for most algorithms that have already been verified with the refinement framework.} of LLVM in Isabelle/HOL, and provide a code generator to actual LLVM text (\S\ref{sec:semantics}). 
On top of this semantics, we build a separation logic and a verification condition generator, which allows convenient reasoning about LLVM programs (\S\ref{sec:vcg}).
Finally, we modify the Sepref tool to target our LLVM semantics instead of Imperative/HOL (\S\ref{sec:auto_ref}). 
This way, we connect the Refinement Framework to our LLVM code generator. This only affects the last refinement step, 
such that most parts of existing verifications can be reused. As a case study, we apply only minor modifications to an existing formalization~\cite{HeLa17} of
the Knuth-Morris-Pratt string search algorithm~\cite{KMP77}, and obtain a verified 
LLVM implementation that is an order of magnitude faster and uses less memory than the originally verified implementation in Standard ML.
The Isabelle theories described in this paper are available at \url{http://www21.in.tum.de/~lammich/isabelle_llvm}.

\section{LLVM Semantics}\label{sec:semantics}
\subsection{State Monad}
The basis of our LLVM semantics is a state-error monad, which we use to conveniently model the preconditions of instructions, their effect 
on memory, as well as arbitrary recursive programs. We define the algebraic data types
\begin{lstlisting}
('a,'s) M = M (run: "'s => ('a,'s) mres")              ('a,'s) mres = NTERM | FAIL | SUCC 'a 's
\end{lstlisting}
An entity of type \q{\is{('a,'s) M}} contains a function \q{\is{run}} that maps a start state of type \q{\is{'s}} to 
a \emph{monad result} that indicates either nontermination, a failure, or a successful execution with a result of type \q{\is{'a}} and a new state.
We define the standard monad combinators:
\begin{lstlisting}
  "return x  = M (%s. SUCC x s)"                                           "get   = M (%s. SUCC s s)"
  "fail         = M (%_. FAIL)"                                                  "set s = M (%_. SUCC () s)"
  "bind m f   = M (%s. case run m s of SUCC x s => run (f x) s | r => r)"
  "assert \<Phi> = if \<Phi> then return () else fail"
\end{lstlisting}
That is, \q{\is{return x}} returns result \q{\is{x}} without changing the state, \q{\is{fail}} aborts the computation,
\q{\is{get}} returns the current state, and \q{\is{set s}} updates the current state.
Finally, \q{\is{bind m f}} first executes \q{\is{m}}, and then \q{\is{f}} with the result of \q{\is{m}}.
If \q{\is{m}} fails or does not terminate, the whole bind fails or does not terminate.  
The derived \q{\is{assert \<Phi>}} combinator can be conveniently used to abort the computation 
if some precondition is violated, e.g., on division by zero.

We use do-notation, i.e.\ \q{\is|do { x<-m; f x }|} is short for \q{\is{bind m (\<lambda>x. f x)}}.
Moreover, we define a flat CCPO\cite{Mark76} on \q{\is{mres}}, with \q{\is{\<bot> := NTERM}}.
We define \q{\is{REC F}} to be the least fixed point of a monotonic function \q{\is{F :: ('a => ('b,'s) M) => 'a => ('b,'s) M}}.
As functions defined using the monad combinators are monotonic by construction~\cite{Kr10}, 
we can define arbitrary recursive computations. The partial function package~\cite{Kr10} supports automation for monotonicity 
proofs and for defining simple recursive functions. Mutual recursion still requires some manual effort, though it could be automated, too.

@@

\subsection{Memory Model}
We use a high-level memory model that does not directly expose the bit-level representation of values, and assumes an infinite supply of memory. 
The memory is modeled as a list of blocks. Each block is either deallocated, or it is a list of values.
A value is a pair of values, a pointer, or an integer. We model memory by the following data types\footnote{We have simplified the presentation a bit. 
The actual implementation defines the concepts memory, block, and value independently, in order to ease future extensions, to, e.g., arbitrary structs instead of pairs.
% . The memory is parameterized over the block type, 
% the block type is parameterized over the value type, and the value type is parameterized over the primitive value type. This allows to 
% exchange parts of the memory model (e.g.\ include arbitrary structures instead of pairs) without breaking the rest of the memory model. 
% What we present here is the instance of our generic memory model that we use for this project.
}:
\begin{lstlisting}
memory = MEMORY (block option list)             block     = val list 
val = PAIR val val | PRIM primval                     primval = PV_INT lint | PV_PTR rptr
\end{lstlisting}
%
Here, the type \q{\is{lint}} is a fixed bit width integer type with a two's complement semantics, as used by LLVM. 
The type \q{\is{rptr}} is either null, or an address. An address is a path through the memory structure to a value:
\begin{lstlisting}
rptr  = NULL | ADDR nat nat (va_dir list)                                 va_dir = PFST | PSND
\end{lstlisting}
An address consists of a \emph{block index}, a \emph{value index}, and a \emph{value address}, which is a list of directions 
to either descend into the first or the second value of a pair.

For the rest of this paper, we will use the state monad with a memory as state. Thus, we define the type \q{\is{'a llM = ('a,memory) M}}.
It is straightforward to define functions \q{\is{load :: rptr => val llM}} and \q{\is{put :: val => rptr => unit llM}} to read/write a value from/to a pointer, 
or fail if the pointer is invalid.
For the actual store function, we check that the structure of the value does not change, i.e. pairs remain pairs, pointers remain pointers, and integers of width $w$ remain integers of width $w$.
\begin{lstlisting}
  store x p = do { y <- load p; assert (vstruct x = vstruct y); put x p }
\end{lstlisting}
%
% 
% 
% Given a pointer \q{\is{p :: rptr}}, it is straightforward to define a lens \q{\is{ptr_L p :: val ==> memory}} that focuses on the pointed to value, 
% or fails if the pointer is invalid (null pointer, pointer to deallocated block, index out of bounds).
% Using this lens, we define a load and store function:
% \begin{lstlisting}
%   load p = use (ptr_L p)
%   store x p = do { y <- use (ptr_L p); assert (vstruct x = vstruct y); ptr_L p := x }
% \end{lstlisting}
% The load function simply returns the pointed-to value. The store function additionally checks that the structure of the value does not change, i.e.,
% pairs remain pairs, pointers remain pointers, and integers of width $w$ remain integers of width $w$. 
% This check is important in order to map our memory model to the actual LLVM memory model, which allocates an amount of memory matching the type of the value.
%
Similarly, we define an allocate and a free function:\\[-2ex]
\begin{minipage}[t]{.46\textwidth}
\begin{lstlisting}
  allocn v n = do {
    blocks <- get; 
    set (blocks@[Some (replicate n v)]);
    return (ADDR |blocks| 0 []) }
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{.52\textwidth}
\begin{lstlisting}
  free (ADDR bi 0 []) = do {
    blocks <- get; 
    assert (bi < |blocks| \and blocks!bi ~= None);
    set (blocks[bi:=None]) }
  free _ = fail
\end{lstlisting}
\end{minipage}
Here, \q{\is{l_1@l_2}} concatenates two lists, \q{\is{|l|}} is the length of list \q{\is{l}}, \q{\is{l!i}} is the $i$th 
element of \q{\is{l}}, and \q{\is{l[i:=x]}} replaces the $i$th element of \q{\is{l}} by \q{\is{x}}.
The allocate function takes an initial value and a block size, appends a new block to the memory, and returns a pointer to the start of the new block (value index $0$, and value address $[]$).
The free function expects a pointer to the start of a block,
checks that this block is not already deallocated, and then deallocates the block by setting it to \q{\is{None}}.

\subsection{Towards a Shallow Embedding}
While we explicitly model values in memory by the type \q{\is{val}}, we model values in registers in a more shallow fashion:
We will identify LLVM registers with Isabelle variables that have a type of the shape \q{\is{T = T \x T | n word | T ptr}}. 
Here, \q{\is{\x}} is Isabelle's product type, \q{\is{n word}} is the $n$ bit word type from Isabelle's word 
library\cprotect\footnote{For convenient notation, we use the type \q{\is{n word}} as if it were a type depending on a variable $n$. 
Isabelle/HOL is not dependently typed. Instead, $n$ is actually a type variable with type-class \q{\is{len}}, which provides a 
function \q{\is{len_of :: 'a::len itself => nat}} to extract the length as a term.}, 
and \q{\is{'a ptr}} is a pointer with an attached phantom type for the value pointed to (\q{\is{'a ptr = PTR rptr}}).
%
For each type \q{\is{'a}} of shape \q{\is{T}}, we define the functions:
\begin{lstlisting}
to_val      :: "'a => val"             struct_of :: "'a itself => vstruct"
from_val :: "val => 'a"             init          :: 'a
\end{lstlisting}
such that
\begin{lstlisting}
"from_val o to_val = id"                "vstruct (to_val x) = (struct_of TYPE('a))"
"to_val init             = zero_initializer (struct_of TYPE('a))"
\end{lstlisting}
Here, \q{\is{TYPE('a) :: 'a itself}} reflects type \q{\is{'a}} into a term.
The functions \q{\is{to_val}} and \q{\is{from_val}} inject a T-shaped type \q{\is{'a}} into a value with structure \q{\is{struct_of TYPE('a)}}.
Moreover, \q{\is{init::'a}} corresponds to the all-zeroes value, i.e., the value where all pointers are null pointers, and all integers are $0$.

\subsection{LLVM Instructions}
In a next step, we define the LLVM instructions. Each LLVM instruction is identified with an Isabelle function. For example, the load instruction
is modeled by:
\begin{lstlisting}
  ll_load :: "'a ptr => 'a llM"
  ll_load (PTR p) = do {
    v <- load p;
    assert (vstruct v = struct_of TYPE('a));
    return (from_val v) }
\end{lstlisting}
It loads a value from the specified pointer, checks that its structure matches the expected type \q{\is{'a}}, and then 
converts the value to \q{\is{'a}}.

For allocation and deallocation, we provide the instructions:
\begin{lstlisting}
  ll_malloc :: 'a itself => n word => 'a ptr llM            ll_free :: "'a ptr => unit llM"
\end{lstlisting}
Note that LLVM does not contain a heap manager. 
Instead, we assume that the generated code will be linked with the C standard library, 
and produce calls to \q{\is{calloc}} and \q{\is{free}}.

We also define
instructions to extract/insert the first/second element of a pair, to offset a pointer, and to advance a pointer to the first/second element.
The code generator maps these instructions to the corresponding LLVM instructions \q{\is{getelementptr}}, \q{\is{insertvalue}}, and \q{\is{extractvalue}}.

The non-memory-related instructions are defined directly on the shallow embedding. For example, we define:
\begin{lstlisting}
  ll_udiv :: "n word => n word => n word"
  ll_udiv a b = do { assert (a ~= 0); return (a div b) }
\end{lstlisting}
where \q{\is{div}} is the unsigned division from the word library.
Note that we use assertions to model undefined behavior in LLVM by failure in our semantics.
Similarly, we define the other integer instructions.

>>>>

\documentclass[a4paper,USenglish,cleveref, autoref]{lipics-v2019}
% \documentclass[a4paper,oribibl,envcountsame]{llncs}
% \pagestyle{plain}
% \usepackage[T1]{fontenc}
% \usepackage[misc]{ifsym}
 \usepackage{amsmath}
 \usepackage{amssymb}
% \usepackage{booktabs}
% \usepackage{bussproofs}
% \usepackage{cite}
% \usepackage{color}
% \usepackage{enumitem}
%\let\vec=\undefined
%\usepackage{mdsymbol}  %%% indirectly needs texlive-generic-extra
 \usepackage{stmaryrd}
% \usepackage{subfig}
% \usepackage{textcomp}
% \usepackage{url}
% \usepackage{xspace}
% \usepackage{mathtools}
% \usepackage{todonotes}
% \usepackage{enumitem}

@@

\usepackage{relsize}

% \usepackage{array}
% \newcolumntype{x}[1]{>{\centering\arraybackslash\hspace{0pt}}b{#1}}
% \usepackage{diagbox}
% \usepackage{siunitx}

@@

% \usepackage[final,tracking=true,kerning=true,spacing=true,stretch=10,shrink=10]{microtype}

%% BEGIN Times font
% \usepackage{mathptmx}
% \usepackage[scaled=.82]{beramono}
% \usepackage[scaled=.86]{helvet}
% \DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
%% END Times font

@@

% \usepackage[
%    a4paper,
%    pdftex,
%    pdftitle={Isabelle LLVM},
%    pdfauthor={Peter Lammich},
%    pdfkeywords={},
%    pdfborder={0 0 0},
%    draft=false,
%    bookmarksnumbered,
%    bookmarks,
%    bookmarksdepth=2,
%    bookmarksopenlevel=2,
%    bookmarksopen]{hyperref}

% \hyphenation{Isa-belle Nieuw-en-huis meta-logic multi-set multi-sets}

\bibliographystyle{plainurl}% the mandatory bibstyle

@@

\ccsdesc[500]{Theory of computation~Program verification}
\ccsdesc[300]{Theory of computation~Logic and verification}
\ccsdesc[300]{Theory of computation~Separation logic}
% \ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Isabelle/HOL,LLVM,Separation Logic,Verification Condition Generator,Code Generation}%TODO mandatory; please add comma-separated list of keywords

@@

\title{Generating Verified LLVM from Isabelle/HOL}
% \titlerunning{Isabelle LLVM}

\author{Peter Lammich}{The University of Manchester}{peter.lammich@manchester.ac.uk}{}{}
\authorrunning {P. Lammich}

% \author{Peter Lammich}
% \authorrunning {P. Lammich}
% \institute{
%   The University of Manchester, England
%   \email{peter.lammich@manchester.ac.uk}
% %   Technische Universit\"at M\"unchen, Munich, Germany \\
% %   \email{lammich@in.tum.de}
% }

@@

\begin{abstract}
We present a framework to generate verified LLVM programs from Isabelle/HOL.
It consists of a code generator that generates LLVM text from 
a simplified fragment of LLVM, shallowly embedded into Isabelle/HOL.
On top, we have developed a separation logic, a verification condition 
generator, and an LLVM backend to the Isabelle Refinement Framework.

As case studies, we have produced verified LLVM implementations of binary search and the 
Knuth-Morris-Pratt string search algorithm. These are one order of magnitude faster than 
the Standard-ML implementations produced with the original Refinement Framework, and on par 
with unverified C implementations. Adoption of the original correctness proofs to the new LLVM backend was straightforward.

The trusted code base of our approach is the shallow embedding of the LLVM fragment 
and the code generator, which is a pretty printer combined with some straightforward compilation steps.

% We formalize the semantics of Isabelle-LLVM, an imperative language shallowly 
% embedded into Isabelle/HOL, designed to be easily translated to actual LLVM text.
% On top of Isabelle-LLVM, we have developed
% \begin{enumerate}
% \item a code generator, to translate from Isabelle-LLVM to actual LLVM text; 
% \item a verification condition generator, to reason about Isabelle-LLVM programs;
% \item a backend for the Isabelle Refinement Framework, to enable a high-level refinement based approach to developing verified LLVM programs.
% \end{enumerate}
% 
% As an example, we adopt an existing formalization of the Knuth-Morris-Pratt string search algorithm to our new framework.
% With only slight changes to the original formalization, we obtain a verified LLVM implementation, which runs faster and uses less memory 
% than the originally verified Standard ML implementation.
% 
% The trusted code base of our approach is the Isabelle-LLVM semantics, which we have engineered for conciseness and readability, 
% and the code generator that translates from Isabelle to LLVM text, which is merely a pretty printer. 
% The other tools that we developed cannot compromise the soundness of the approach, as they only generate theorems that are proved via Isabelle's kernel.
\end{abstract}

\section{Introduction}\label{sec:introduction}

% Current code generator technology in Isabelle and other interactive theorem provers is limited to emit high-level functional target 
% languages, which are inherently less efficient than low-level imperative languages such as C or LLVM~\cite{LLVM-manual}.
% This work presents a framework to generate verified LLVM programs from Isabelle/HOL~\cite{NPW02,NiKl14}.
% 
% We developed the Isabelle-LLVM language, an imperative language shallowly embedded into Isabelle/HOL.
% Isabelle-LLVM is designed to be easily translated into actual LLVM text, which is done by our code generator.
% On top of Isabelle/LLVM we build a separation logic and a verification condition generator, which enables us to verify
% basic data structures and algorithms. To simplify verification of more complex algorithms, we provide a stepwise refinement 
% approach, based on the Isabelle Refinement Framework~\cite{LaTu12,La13}.

The Isabelle Refinement Framework~\cite{LaTu12,La13,La15} features a stepwise refinement approach to verified algorithms, using the Isabelle/HOL theorem prover~\cite{NPW02,NiKl14}.
It has been successfully applied to verify many algorithms and software systems, 
among them LTL and timed automata model checkers~\cite{ELNN13,BrLa18,WiLa18}, network flow algorithms~\cite{LaSe17,LaSe16}, a SAT-solver certification tool~\cite{La17_CADE,La17_SAT}, and even a SAT solver~\cite{FBL18}.
Using Isabelle/HOL's code generator~\cite{HKKN13}, the verified algorithms can be extracted to functional languages like Haskell or Standard ML.
However, the code generator only provides partial correctness guarantees, i.e., termination of the generated code cannot be proved.
Moreover, the generated code is typically slower than the same algorithms implemented in C or Java.

The original Refinement Framework~\cite{LaTu12,La13} could only generate purely functional code. 
The first remedy to the performance problem was to introduce array data structures that behave like functional 
lists on the surface, but are implemented by destructively updated arrays behind the scenes, similar to Haskell's now deprecated DiffArray.
While this gained some performance, the array implementation itself was not verified, such that we had to trust its correctness. 
Moreover, an array access still required a significant amount of overhead compared to a simple pointer dereference in C.

The next step towards more efficient verified implementations was the Sepref tool~\cite{La15}. 
It generates code for Imperative HOL~\cite{BKHEM08}, which provides a heap monad inside Isabelle/HOL, 
and a code generator extension to generate code that uses the stateful arrays provided by ML, or the heap monad of Haskell.
The Sepref tool performs automatic data refinement from abstract data types like maps or sets to concrete implementations like hash tables, which 
can be placed on the heap and destructively updated. 
Moreover, it provides tools~\cite{La16} to assist in the definition of new data structures, exploiting `free theorems'~\cite{Wad89} that it 
obtains from parametricity properties of the abstract data types.
%
Using Imperative HOL as backend, we gained some additional performance: For example, the GRAT tool~\cite{La17_CADE,La17_SAT} provides a 
verified checker for UNSAT certificates in the DRAT format~\cite{WHH14}. It is faster than the unverified state-of-the art 
checker \textsc{drat-trim}~\cite{WHH14}, which is written in C. However, the GRAT tool spends most of its run time in an unverified 
certificate preprocessor. Nevertheless, optimizing the verified part of the code is important: The very same technique was also implemented in Coq,
using purely functional data structures~\cite{CMS17,CHHKS17}. There, the verified code was actually the 
bottleneck\footnote{Later, the checker was rewritten in ACL2, also using imperative data structures~\cite{CHHKS17,HHKW17}}.

This paper presents a next step towards efficient verified algorithms: A refinement framework to generate verified code in LLVM intermediate representation~\cite{LLVM-manual} 
with total correctness guarantees. LLVM is an imperative intermediate language with a powerful and well-tested optimizing compiler.
We first formalize the semantics of Isabelle-LLVM, a simple imperative language shallowly embedded into Isabelle/HOL, 
and designed to be easily translated to actual LLVM text (\S\ref{sec:semantics}). 
On top of Isabelle-LLVM, we build a separation logic and a verification condition generator, which allows convenient reasoning about Isabelle-LLVM programs (\S\ref{sec:vcg}).
Finally, we modify the Sepref tool to target Isabelle-LLVM instead of Imperative/HOL (\S\ref{sec:auto_ref}), connecting 
the Refinement Framework to our LLVM code generator. This only affects the last refinement step, 
such that most parts of existing verifications can be reused. 
As case studies (\S\ref{sec:casestudies}), we verify a binary search algorithm and adopt an existing formalization~\cite{HeLa17} of
the Knuth-Morris-Pratt string search algorithm~\cite{KMP77}. The resulting LLVM code is significantly
faster than the corresponding Standard-ML code and on par with unverified C implementations.
%
The Isabelle theories described in this paper are available at \url{http://www21.in.tum.de/~lammich/isabelle_llvm}.

\section{Isabelle-LLVM}\label{sec:semantics}
\subsection{State Monad}
The basis of Isabelle-LLVM is a state-error monad, which we use to conveniently model the preconditions of instructions, their effect 
on memory, as well as arbitrary recursive programs. We define the algebraic data types:
\begin{lstlisting}
('a,'s) M = M (run: "'s => ('a,'s) mres")        ('a,'s) mres = NTERM | FAIL | SUCC  'a  's
\end{lstlisting}
An entity of type \q{\is{('a,'s) M}} contains a function \q{\is{run}} that maps a start state of type \q{\is{'s}} to 
a \emph{monad result} that indicates either nontermination, a failure, or a successful execution with a result of type \q{\is{'a}} and a new state.
We define the standard monad combinators:
\begin{lstlisting}
  "return x  = M (%s. SUCC x s)"                                get  = M (%s. SUCC s s)
  "fail       = M (%_. FAIL)"                                     set s = M (%_. SUCC () s)
  "bind m f  = M (%s. case run m s of SUCC x s => run (f x) s | r => r)"
  "assert \<Phi> = if \<Phi> then return () else fail"
\end{lstlisting}
That is, \q{\is{return x}} returns result \q{\is{x}} without changing the state, \q{\is{fail}} aborts the computation,
\q{\is{get}} returns the current state, and \q{\is{set s}} updates the current state.
Finally, \q{\is{bind m f}} first executes \q{\is{m}}, and then \q{\is{f}} with the result of \q{\is{m}}.
If \q{\is{m}} fails or does not terminate, the whole bind fails or does not terminate.  
The derived \q{\is{assert \<Phi>}} combinator can be conveniently used to abort the computation 
if some precondition is violated, e.g., on division by zero.

We use do-notation, i.e.\ \q{\is|do { x<-m; f x }|} is short for \q{\is{bind m (\<lambda>x. f x)}}.
Moreover, we define a flat chain complete partial order~\cite{Mark76} on \q{\is{mres}}, with \q{\is{\<bot> := NTERM}}.
For a monotonic function \q{\is{F :: ('a => ('b,'s) M) => 'a => ('b,'s) M}}, \q{\is{REC F}} is the least fixed point.
As functions defined using the monad combinators are monotonic by construction~\cite{Kr10}, 
we can define arbitrary recursive computations. The partial function package~\cite{Kr10} supports automation for monotonicity 
proofs and for defining simple recursive functions. Mutual recursion still requires some manual effort, though it could be automated, too.

@@

\subsection{Memory Model}
We use a high-level memory model that does not directly expose the bit-level representation of values and assumes an infinite supply of memory. 
The memory is modeled as a list of blocks. Each block is either deallocated, or it is a list of values.
A value is a pair of values, a pointer, or an integer. We model memory by the following data types\footnote{We have slightly simplified the presentation. 
The actual implementation defines the concepts memory, block, and value in a modular fashion, in order to ease future extensions.
% . The memory is parameterized over the block type, 
% the block type is parameterized over the value type, and the value type is parameterized over the primitive value type. This allows to 
% exchange parts of the memory model (e.g.\ include arbitrary structures instead of pairs) without breaking the rest of the memory model. 
% What we present here is the instance of our generic memory model that we use for this project.
}:
\begin{lstlisting}
memory = MEMORY (block list)                 block     = val list option
val = PAIR val val | PRIM primval              primval = PV_INT lint | PV_PTR rptr
\end{lstlisting}
%
Here, the type \q{\is{lint}} is a fixed bit width word type with a two's complement semantics, as used by LLVM. 
The type \q{\is{rptr}} is either null or an address. An address is a path through the memory structure to a value:
\begin{lstlisting}
rptr  = NULL | ADDR nat nat (va_dir list)                      va_dir = PFST | PSND
\end{lstlisting}
An address consists of a \emph{block index}, a \emph{value index}, and a \emph{value address}, which is a list of directions 
to either descend into the first or the second value of a pair.

For the rest of this paper, we will use the state monad with a memory as state. Thus, we define the type \q{\is{'a llM = ('a,memory) M}}.
It is straightforward to define functions \q{\is{load :: rptr => val llM}} and \q{\is{put :: val => rptr => unit llM}} to read/write a value from/to a pointer, 
or fail if the pointer is invalid.
For the actual store function, we check that the structure of the value does not change, i.e. pairs remain pairs, pointers remain pointers, and words of width $w$ remain words of width $w$:
\begin{lstlisting}
  store x p = do { y <- load p; assert (vstruct x = vstruct y); put x p }
  $\text{where}$
  vstruct (PAIR a b) = VS_PAIR (vstruct a) (vstruct b)
  vstruct (PRIM (PV_PTR _)) = VS_PTR
  vstruct (PRIM (PV_INT w)) = VS_INT (width w) 
\end{lstlisting}

%
% 
% 
% Given a pointer \q{\is{p :: rptr}}, it is straightforward to define a lens \q{\is{ptr_L p :: val ==> memory}} that focuses on the pointed to value, 
% or fails if the pointer is invalid (null pointer, pointer to deallocated block, index out of bounds).
% Using this lens, we define a load and store function:
% \begin{lstlisting}
%   load p = use (ptr_L p)
%   store x p = do { y <- use (ptr_L p); assert (vstruct x = vstruct y); ptr_L p := x }
% \end{lstlisting}
% The load function simply returns the pointed-to value. The store function additionally checks that the structure of the value does not change, i.e.,
% pairs remain pairs, pointers remain pointers, and integers of width $w$ remain integers of width $w$. 
% This check is important in order to map our memory model to the actual LLVM memory model, which allocates an amount of memory matching the type of the value.
%
Similarly, we define an allocate and a free function:\\
\begin{minipage}[t]{.46\textwidth}
\begin{lstlisting}
  allocn v n = do {
    blocks <- get; 
    set (blocks@[Some (replicate n v)]);
    return (ADDR |blocks| 0 []) }
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{.52\textwidth}
\begin{lstlisting}
  free (ADDR bi 0 []) = do {
    blocks <- get; 
    assert (bi < |blocks| \and blocks!bi ~= None);
    set (blocks[bi:=None]) }
  free _ = fail
\end{lstlisting}
\end{minipage}
Here, \q{\is{l_1@l_2}} concatenates two lists, \q{\is{|l|}} is the length of list \q{\is{l}}, \q{\is{l!i}} is the $i$th 
element of \q{\is{l}}, and \q{\is{l[i:=x]}} replaces the $i$th element of \q{\is{l}} by \q{\is{x}}.
The allocate function takes an initial value and a block size, appends a new block to the memory, and returns a pointer to the start of the new block (value index $0$, and value address $[]$).
The free function expects a pointer to the start of a block,
checks that this block is not already deallocated, and then deallocates the block by setting it to \q{\is{None}}.

\subsection{Towards a Shallow Embedding}
While we explicitly model values in memory by the type \q{\is{val}}, we model values in registers in a more shallow fashion:
We identify LLVM registers with Isabelle variables that have a type of the shape \q{\is{T = T \x T | n word | T ptr}}. 
Here, \q{\is{\x}} is Isabelle's product type, \q{\is{n word}} is the $n$ bit word type from Isabelle's word 
library\cprotect\footnote{For convenient notation, we use the type \q{\is{n word}} as if it were a type depending on a variable $n$. 
Isabelle/HOL is not dependently typed. Instead, $n$ is actually a type variable with type-class \q{\is{len}}, which provides a 
function \q{\is{len_of :: 'a::len itself => nat}} to extract the length as a term.}, 
and \q{\is{'a ptr}} is a pointer with an attached phantom type for the value pointed to (\q{\is{'a ptr = PTR rptr}}).
%
For each type \q{\is{'a}} of shape \q{\is{T}}, we define the functions:
\begin{lstlisting}
to_val     :: "'a => val"            struct_of    :: "'a itself => vstruct"
from_val :: "val => 'a"             init          :: 'a
\end{lstlisting}
such that
\begin{lstlisting}
"from_val o to_val = id"                "vstruct (to_val x) = (struct_of TYPE('a))"
"to_val init = zero_initializer (struct_of TYPE('a))"
\end{lstlisting}
Here, \q{\is{TYPE('a) :: 'a itself}} reflects type \q{\is{'a}} into a term.
The functions \q{\is{to_val}} and \q{\is{from_val}} inject a T-shaped type \q{\is{'a}} into a value with structure \q{\is{struct_of TYPE('a)}}.
Moreover, \q{\is{init::'a}} corresponds to the all-zeroes value, i.e., the value where all pointers are null pointers, and all integers are $0$.

\subsection{Instructions}
In a next step, we define the instructions of Isabelle-LLVM. Each instruction is identified with an Isabelle constant. For example, the load instruction
is modeled by:
\begin{lstlisting}
  ll_load :: "'a ptr => 'a llM"
  ll_load (PTR p) = do {
    v <- load p;
    assert (vstruct v = struct_of TYPE('a));
    return (from_val v) }
\end{lstlisting}
It loads a value from the specified pointer, checks that its structure matches the expected type \q{\is{'a}}, and then 
converts the value to \q{\is{'a}}.

For allocation and deallocation, we provide the instructions:
\begin{lstlisting}
  ll_malloc :: 'a itself => n word => 'a ptr llM            ll_free :: "'a ptr => unit llM"
\end{lstlisting}
Note that LLVM does not contain a heap manager. 
Instead, we assume that the generated code will be linked with the C standard library, 
and let the code generator produce calls to \q{\is{calloc}} and \q{\is{free}}.
%
We also define instructions to access the elements of a pair, to offset a pointer, and to advance a pointer into a pair.
The code generator maps these instructions to the corresponding LLVM instructions \q{\is{getelementptr}}, \q{\is{insertvalue}}, and \q{\is{extractvalue}}.
